---
title: "Michael Jordan vs LeBron James"
author: "Ryan Gavin"
date: "6/9/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

### Executive Summary

As an example of practice with the `plot_ly` package, we'll take a look at some of the stats for **Michael Jordan** and **LeBron James** over their careers.

### Data Processing

We'll begin by loading the necessary libraries. Then we read in the data$^*$ from `.csv` files and tidy it up.

```{r libraries, echo=FALSE}
require(ggplot2)
require(plotly)
require(dplyr)
```

```{r data}
mj_pg <- tbl_df(read.csv("MJ_PerGame.csv",stringsAsFactors = FALSE))
lj_pg <- tbl_df(read.csv("LJ_PerGame.csv",stringsAsFactors = FALSE))

mj_pg$Season[1:dim(mj_pg)[1]] <- c(1:dim(mj_pg)[1])
lj_pg$Season[1:dim(lj_pg)[1]] <- c(1:dim(lj_pg)[1])

mj_pg <- mutate(mj_pg,"Lg" = NULL, "Pos" = NULL)
lj_pg <- mutate(lj_pg,"Lg" = NULL, "Pos" = NULL)

mj_pg$Season <- as.numeric(mj_pg$Season)
lj_pg$Season <- as.numeric(lj_pg$Season)

mj_pg <- mutate(mj_pg, "Player" = "Jordan")
lj_pg <- mutate(lj_pg, "Player" = "James")

pg <- rbind(mj_pg,lj_pg)
```

### Data Analysis

The following plot shows various statistics for the players Michael Jordan and LeBron James over their respective careers. Statistics (per season) included are  
- Games Played  
- Minutes Played per Game  
- Average Points per Game  
- Average Offensive Rebounds per Game  
- Average Defensive Rebounds per Game  
- Average Total Rebounds per Game  
- Average 2-Point Field Goals per Game (Made, Attempted, Percentage)  
- Average 3-Point Field Goals per Game (Made, Attempted, Percentage)  
- Average Total Field Goals per Game (Made, Attempted, Percentage)  

Use the buttons on the plot to navigate through the various statistics.

```{r analysis, echo=FALSE, fig.height=7.5, fig.width=10}
updatemenus <- list(
      list(
            type="buttons",
            buttons = list(

                  # #1                  
                  list(
                        label = "Games Played",
                        method = "update",
                        args = list(list(visible=c(rep(TRUE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2))),
                                    list(title = "M. Jordan vs L. James: Games Played In",
                                         yaxis=list(title="Games Played")))
                        ),

                  # #2
                  list(
                        label = "Minutes Played",
                        method = "update",
                        args = list(list(visible=c(rep(FALSE,4),
                                                   rep(TRUE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2))),
                                    list(title = "M. Jordan vs L. James: Avg Minutes Played Per Game",
                                         yaxis=list(title="Minutes Played/Game")))
                        ),

                  # #3
                  list(
                        label = "Points",
                        method = "update",
                        args = list(list(visible=c(rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(TRUE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2))),
                                    list(title = "M. Jordan vs L. James: Avg Points per Game",
                                         yaxis=list(title="Points/Game")))
                        ),

                  # #4
                  list(
                        label = "Off Rebounds",
                        method = "update",
                        args = list(list(visible=c(rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(TRUE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2))),
                                    list(title="M. Jordan vs L. James: Offensive Rebounds per Game",
                                         yaxis=list(title="Off Rebounds/Game")))
                        ),
                  # #5
                  list(
                        label = "Def Rebounds",
                        method = "update",
                        args = list(list(visible=c(rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(TRUE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2))),
                                    list(title="M. Jordan vs L. James: Defensive Rebounds per Game",
                                         yaxis=list(title="Def Rebounds/Game")))
                        ),
                  # #6
                  list(
                        label = "Total Rebounds",
                        method = "update",
                        args = list(list(visible=c(rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(TRUE,4),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2))),
                                    list(title="M. Jordan vs L. James: Total Rebounds per Game",
                                         yaxis=list(title="Total Rebounds/Game")))
                        ),
                  
                  # #7, 8, 9
                  list(
                        label = "2-PT Field Goals",
                        method = "update",
                        args = list(list(visible=c(rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(TRUE,2),
                                                   rep(TRUE,2),
                                                   rep(TRUE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2))),
                                    list(title="M. Jordan vs L. James: 2-PT Field Goals per Game",
                                         yaxis=list(side="left",title="2-PT FG Made and Attempted/Game"),
                                         yaxis2=list(side="right",overlaying="y",range=c(10,70),
                                                     title="2-PT FG Percentage/Game")))
                        ),
                  
                  # #10, 11, 12
                  list(
                        label = "3-PT Field Goals",
                        method = "update",
                        args = list(list(visible=c(rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(TRUE,2),
                                                   rep(TRUE,2),
                                                   rep(TRUE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2))),
                                    list(title="M. Jordan vs L. James: 3-PT Field Goals per Game",
                                         yaxis=list(side="left",title="3-PT FG Made and Attempted/Game"),
                                         yaxis2=list(side="right",overlaying="y",range=c(10,70),
                                                     title="3-PT FG Percentage/Game")))
                        ),
                  
                  # #13, 14, 15
                  list(
                        label = "Total Field Goals",
                        method = "update",
                        args = list(list(visible=c(rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,4),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(FALSE,2),
                                                   rep(TRUE,2),
                                                   rep(TRUE,2),
                                                   rep(TRUE,2))),
                                    list(title="M. Jordan vs L. James: Total Field Goals per Game",
                                         yaxis=list(side="left",title="Total FG Made and Attempted/Game"),
                                         yaxis2=list(side="right",overlaying="y",range=c(10,70),
                                                     title="Total FG Percentage/Game")))
                        )
                  )
            )
      )


p <- plot_ly(data = pg) %>%
      #1
      add_lines(x=~Season,y=~G,
                type="scatter",mode="lines+markers",
                marker=list(size=10,opacity=0.45),linetype=~Player,color=~Tm) %>%
      #2
      add_lines(x=~Season,y=~MP,
                type="scatter",mode="lines+markers",
                marker=list(size=10,opacity=0.45),linetype=~Player,color=~Tm,visible=FALSE) %>%
      #3
      add_lines(x=~Season,y=~PTS,
                type="scatter",mode="lines+markers",
                marker=list(size=10,opacity=0.45),linetype=~Player,color=~Tm,visible=FALSE) %>%
      #4
      add_lines(x=~Season,y=~ORB,
                type="scatter",mode="lines+markers",
                marker=list(size=10,opacity=0.45),linetype=~Player,color=~Tm,visible=FALSE) %>%
      #5
      add_lines(x=~Season,y=~DRB,
                type="scatter",mode="lines+markers",
                marker=list(size=10,opacity=0.45),linetype=~Player,color=~Tm,visible=FALSE) %>%
      #6
      add_lines(x=~Season,y=~TRB,
                type="scatter",mode="lines+markers",
                marker=list(size=10,opacity=0.45),linetype=~Player,color=~Tm,visible=FALSE) %>%
      #7, 8, 9
      add_trace(x=~Season,y=~X2P,type="bar",name="Made 2-PT FG",
                color=~Player,visible=FALSE) %>%
      add_trace(x=~Season,y=~X2PA,type="bar",name="Attempted 2-PT FG",
                color=~Player,visible=FALSE) %>%
      add_lines(x=~Season,y=~X2P.*100,type="scatter",name="2-PT FG %",mode="lines",
                color=I("black"),linetype=~Player,yaxis="y2",visible=FALSE) %>%
      
      #10, 11, 12
      add_trace(x=~Season,y=~X3P,type="bar",name="Made 3-PT FG",
                color=~Player,visible=FALSE) %>%
      add_trace(x=~Season,y=~X3PA,type="bar",name="Attempted 3-PT FG",
                color=~Player,visible=FALSE) %>%
      add_lines(x=~Season,y=~X3P.*100,type="scatter",name="3-PT FG %",mode="lines",
                color=I("black"),linetype=~Player,yaxis="y2",visible=FALSE) %>%
      
      #13, 14, 15
      add_trace(x=~Season,y=~FG,type="bar",name="Made Total FG",
                color=~Player,visible=FALSE) %>%
      add_trace(x=~Season,y=~FGA,type="bar",name="Attempted Total FG",
                color=~Player,visible=FALSE) %>%
      add_lines(x=~Season,y=~FG.*100,type="scatter",name="Total FG %",mode="lines",
                color=I("black"),linetype=~Player,yaxis="y2",visible=FALSE) %>%
      
      layout(title="Michael Jordan vs LeBron James", 
             yaxis=list(title="Games Played"),
             updatemenus = updatemenus)

p
```

-------------

$^*$ Data was taken from the website [Sports Reference](http://www.sports-reference.com).
